---
title: "Study-size Adjusted Percentages"
author: "Mike Stackhouse"
date: "9/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source('podr_connections.R')
library(tidyverse)
library(Tplyr)
library(knitr)
library(huxtable)
```

Pooled studies present a challenge in presenting apporopriately represented proportions when looking at subject totals in tables like Adverse Events. This document will explore proper calculation of Study-size adjusted percentages, as recommended. For more information on this problem, review [this YouTube video](https://www.youtube.com/watch?v=GGU6-Pmhq-g).

## Recommended weighting scheme

The weighting scheme recommended by PHUSE takes the proportion of subjects in an individual pooled study, within a treatment group compared to the total number of subjected in the pooled studies within that treatment group overall. This weight is then applied when aggregating the incidence across studies to identify the total percentage. 

## Mock some data

We don't have a pooled study available currently, so let's mock some up by randomly sampling subjects from the CDISC pilot into different 'studies'.

```{r get data}
# Note that you will need your own PODR account for this to work for you
connect_podr('phuse_p7r82p0vul')

# Read ADSL and ADAE
adae <- read_podr('adae', libname='cdisc_pilot_adam')
adsl <- read_podr('adsl', libname='cdisc_pilot_adam')
```
With the data loaded, let's create some new ADSL datasets.

```{r mock adsl}
set.seed(1234)
# Get the set of subjects
all_subs <- adsl$USUBJID
# Sample 20 for the first group
study1_subs <- sample(all_subs, 20)
# Remove those subjects from the pool of available subjects
all_subs <- setdiff(all_subs, study1_subs)
# Repeat for study 2
study2_subs <- sample(all_subs, 30)
# Remove from total set
all_subs <- setdiff(all_subs, study2_subs)
# Repeat for study 3
study3_subs <- sample(all_subs, 150)

# Filter each of the ADSLs down
adsl1 <- adsl %>% 
  filter(USUBJID %in% study1_subs) 

adsl2 <- adsl %>% 
  filter(USUBJID %in% study2_subs) %>% 
  mutate(STUDYID = "CDISCPILOT02")

adsl3 <- adsl %>% 
  filter(USUBJID %in% study3_subs) %>% 
  mutate(STUDYID = "CDISCPILOT03")

# Combine data together
adsl_pooled <- bind_rows(adsl1, adsl2, adsl3)

# Do some clean up
rm(all_subs, study1_subs, study2_subs, study3_subs)

```

Great - we have 3 different ADSLs. Let's mock up the matching ADAE datasets

```{r mock adae}
# Merge to subset ADAE to just subjects in the associated ADSL dataset
adae1 <- adsl1 %>% 
  select(STUDYID, USUBJID) %>% 
  inner_join(adae, by='USUBJID') %>% 
  mutate(STUDYID = STUDYID.x) %>% 
  select(-STUDYID.x, -STUDYID.y)

adae2 <- adsl2 %>% 
  select(STUDYID, USUBJID) %>% 
  inner_join(adae, by='USUBJID') %>% 
  mutate(STUDYID = STUDYID.x) %>% 
  select(-STUDYID.x, -STUDYID.y)

adae3 <- adsl3 %>% 
  select(STUDYID, USUBJID) %>% 
  inner_join(adae, by='USUBJID') %>% 
  mutate(STUDYID = STUDYID.x) %>% 
  select(-STUDYID.x, -STUDYID.y)

# Combine data together
adae_pooled <- bind_rows(adae1, adae2, adae3)

# Clean up
rm(adsl1, adsl2, adsl3, adae1, adae2, adae3)

```

Awesome - now we have some pooled datasets to play with. Like CDISC data, subjects are marked in their respective studies within the STUDYID variable. 

## Deriving Weights

The next step is to derive weights. In order to derive weights, we need two pieces of information:

- The total number of subjects in each study
- The total number of subjects overall

This differs from the crude study-size adjustment approach, which would weight by treatment arm.

Simple enough - let's go.

```{r study size weights}
# Start with overall count
total_subs <- adsl_pooled %>%
  summarize(total_subs = n(), .groups='drop') %>% 
  # Magrittr is more than a pipe (%>%)! Look at the help
  # documentation for magrittr::extract. There are a number of nifty
  # tools for working in a tidy context
  extract2('total_subs')

# Now per study
total_study_subs <- adsl_pooled %>% 
  group_by(STUDYID) %>% 
  summarize(total_study_subs = n(), .groups='drop') %>% 
  # Instead of merging, since I have the distinct count 
  # above I can just assign it here
  mutate(total_subs = total_subs)

# Subjects in an ARM
total_study_arm <- adsl_pooled %>% 
  group_by(STUDYID, TRT01P) %>% 
  summarize(total_study_arm = n(), .groups='drop') 

# Get denom dataset assembled
denoms <- total_study_subs %>% 
  full_join(total_study_arm, by='STUDYID') %>% 
  mutate(weight = total_study_subs / total_subs)

rm(total_subs, total_study_subs)

kable(head(denoms))
```

Wonderful - we have some really useful information now - we have the denominator to use when calculating the incidence per study, and then the weight to use when aggregating the study-size adjusted percentage. Now we can go ahead to the next step - counting! 

## Counting Adverse Events

Next we need to count our adverse events. This is the first step before we can additionally derive percentages. Based on the summaries we're deriving, we must do a few other things as well:

- Group by STUDYID, which you typically wouldn't do
- Group by Treatment Group
- Calculate _distinct_ counts by USUBJID - to get the study-size adjusted percentage we need to know the proportion of subjects who experienced an AE, not the number of AEs experienced. 

For simplicity sake, we're just going to summarize preferred term.

```{r count AEs}
counts_by_study <- adae_pooled %>% 
  group_by(STUDYID, TRTA, AEDECOD) %>% 
  distinct(USUBJID) %>%
  summarize(counts = n(), .groups='drop')

kable(head(counts_by_study))
```

Great - Now let's incorporate the denominators and the weights that we derived earlier. This will set us up to perform the final calculation of study-size adjusted percentages.

```{r get weights}
counts_and_weight <- counts_by_study %>% 
  # Using dplyr joins, you don't have to have matching variable names!
  # Instead, you can actually specify the by variables using a named character
  # vector (i.e. using c()). Below you can see that counts_by_study has the 
  # variable TRTA and denoms has TRTP - I instruct dplyr to merge on these
  # two variables using 'left_variable' = 'right_variable'
  left_join(denoms, by=c('STUDYID', 'TRTA' = 'TRT01P'))

kable(head(counts_and_weight))
```

# Deriving Study-size adjusted percentage

So let's think about what has to happen next. 

- The percentage in each study, by treatment, by adverse event needs to be calculated
- Those percentages should be adjusted by the weights we calculated earlier 
- The percentages for each adverse event by study by treatment should be aggregated.

There are a few steps, but the overall process is not very complicated. Let's go through step by step.

```{r adjusted percentage calculation}
adjusted_pcts <- counts_and_weight %>% 
  # Calculate percent incidence of AEs weighted by study and treatment group
  mutate(weighted_pct_by_study = weight * (counts / total_study_arm)) %>% 
  # Group by treatment group and adverse event and sum 
  # now we're aggregating the incidence of the AEs together within a treatment group
  group_by(TRTA, AEDECOD) %>% 
  summarize(adj_pct_n = sum(weighted_pct_by_study), .groups='drop') %>% 
  # Lastly, let's do a little string formatting to make it pretty
  mutate(adj_pct = format(round(adj_pct_n * 100, digits=1), digits=1, width=4, nsmall=1)) %>% 
  # And grab just the variables we want.
  select(TRTA, AEDECOD, adj_pct)

kable(head(adjusted_pcts))
```

Wonderful - we have one last step of pivoting these data for presentation in a final output, and filling in the empty values.

```{r final pivot}
adjusted_pcts_t <- adjusted_pcts %>% 
  # Pivot wider goes from long to wide - so here I'm keeping AEDECOD as is, making
  # my variable names from TRTA, pulling values from adj_pct - so the TRTA values
  # become columns and the associated data from adj_pct fills those columns. 
  # And finally to make my column names useful I add a prefix of adj_pct_. 
  pivot_wider(id_cols = AEDECOD, names_from=TRTA, values_from = adj_pct, names_prefix='adj_pct_') %>% 
  # dplyr has a number of useful mutate variations that have different applications.
  # Instead of applying the same function to all 3 of the columns created in the block above,
  # here I'm saying for columns 2 through 4, use replace_na and the `replace` argument should 
  # be set to ' 0.0'. View the help page for dplyr::mutate_at for more examples.
  mutate_at(2:4, replace_na, replace=' 0.0')

kable(head(adjusted_pcts_t))
```

## Incorporating Into a Final Table

Everything above calculates the study-size adjusted percentage - but we need the rest of the table too. Let's use [Tplyr](https://atorus-research.github.io/Tplyr/) to get the rest of the data we need. 

```{r build table}
# Create the Tplyr table and specify the treatment variable
adae_table <- tplyr_table(adae_pooled, TRTA) %>% 
  # Set the population dataset
  set_pop_data(adsl_pooled) %>%
  # Set the population treatment variable
  set_pop_treat_var(TRT01P) %>% 
  # Add a new layer to the table
  add_layer(
    # Create a count summary on AEDECOD
    group_count(AEDECOD) %>% 
      # Set counts distinct by USUBJID
      set_distinct_by(USUBJID) %>% 
      # Set the order method for the results 
      # to by count
      set_order_count_method('bycount') %>% 
      # Use the distinct subject counts for the order
      set_result_order_var(distinct_n) %>% 
      # Pull the order from the Xanomeline High Dose variable
      set_ordering_cols('Xanomeline High Dose')
  ) 

adae_table_built <- adae_table %>% 
  # Crunch the numbers
  build()

kable(head(adae_table_built))
```

Great - now we have the table data we need. Let's get the adjusted percentages in and do some styling!

```{r merge adjusted percent}
final <- adae_table_built %>% 
  # Join in the adjusted percentages
  left_join(adjusted_pcts_t, by = c('row_label1' = 'AEDECOD')) %>% 
  # Sort by descending incidence from the Xanomeline High Dose group, and 
  # then alphabetically be preferred term
  arrange(desc(ord_layer_1), row_label1) %>% 
  # Select off the variables in the order I want them
  select(row_label1, var1_Placebo, adj_pct_Placebo, `var1_Xanomeline Low Dose`, 
         `adj_pct_Xanomeline Low Dose`, `var1_Xanomeline High Dose`, 
         `adj_pct_Xanomeline High Dose`) %>% 
  # Add a nested set of column headers using Tplyr::add_column_headers. Using 
  # the adae_table object from above, the overall N counts are auto replaced 
  # for the **Treatment Group** strings below
  add_column_headers(
    paste0('Preferred Term | Placebo (N=**Placebo**) {n(%) | Study Size-Adjusted %} | ' ,
           'Xanomeline Low Dose (N=**Xanomeline Low Dose**) {n(%) | Study Size-Adjusted %} |',
           'Xanomeline High Dose (N=**Xanomeline High Dose**) {n(%) | Study Size-Adjusted %}'),
     header_n = header_n(adae_table)
  )

```

And finally - let's do some styling using the library Huxtable

```{r styling}
ht <- as_hux(final, add_colnames=FALSE) %>%
  set_bold(1:2, 1:ncol(final), TRUE) %>% # bold the first row
  set_align(1:2, 1:ncol(final), 'center') %>% # Center align the first row 
  set_align(3:nrow(final), 2:ncol(final), 'center') %>% # Center align the results
  set_valign(1:2, 1:ncol(final), 'bottom') %>% # Bottom align the first row
  set_bottom_border(2, 1:ncol(final), 1) %>% # Put a border under the first row
  set_width(1.1) %>% # Set the table width
  set_col_width(c(.3, rep(.7/6, 6))) %>%  # Set the column widths
  merge_cells(1, 2:3) %>% # Merge placebo spanning header
  merge_cells(1, 4:5) %>% # Merge low dose spanning header
  merge_cells(1, 6:7) %>%  # Merge high dose spanning header
  set_background_color(seq(3, nrow(final), 2), everywhere, "grey95")
ht
```

```{r close_connection, include=FALSE}
DBI::dbDisconnect(getOption('podr_connection'))
```